---
title: "Reshaping Data with Tidyr"
format: 
  html:
    embed-resources: true
editor: visual
---

Reshaping Data

[TidyR website](https://tidyr.tidyverse.org/)

[TidyR CRAN package site](https://cran.r-project.org/web/packages/tidyr/index.html)


[Cheatsheet](https://rstudio.github.io/cheatsheets/tidyr.pdf)




# Pivoting

## pivot_longer()

```{r}
#| label: seetup
#| echo: false
#| warning: false
#| message: false

library(tidyr)
library(dplyr)
library(readr)



```

### Religion

Example from last class:
```{r}
relig_income

```



This dataset contains **three variables**:

  - **religion**, stored in the rows,
  - **income** spread across the column names, and
  - **count** stored in the cell values.
  
  
#### Use pivot_longer():

  - The first argument is the dataset to reshape, relig_income.

  - cols describes which columns need to be reshaped. In this case, it’s every column apart from religion.

  - names_to gives the name of the variable that will be created from the data stored in the column names, i.e. income.

  - values_to gives the name of the variable that will be created from the data stored in the cell value, i.e. count.

  - Neither the names_to nor the values_to column exists in relig_income, so we provide them as strings surrounded by quotes.


```{r}
#| label: religion example
#| echo: true

relig_income %>% 
  pivot_longer(
    cols = !religion, 
    names_to = "income", 
    values_to = "count"
  )

```



### Billboard: Numeric data in column names


```{r}
#| label: billboard

billboard

```


  - The names become a variable called week

  - The values to become a variable called rank

  - Use values_drop_na to drop rows with  missing values which are created with 
a song drops off the charts in less than 76 weeks.

```{r}
#| label: billboard pivot 1

billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )

```

Improved

  - Show how long each song stayed in the charts by converting the week variable to an integer
       - use names_prefix to strips off the wk prefix, 
       - use names_transform to convert week into an integer
       
```{r}
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    names_prefix = "wk",
    names_transform = as.integer,
    values_to = "rank",
    values_drop_na = TRUE,
  )

```



### WHO

```{r}
#| label: WHO

who


```

   - country, iso2, iso3, and year are already variables. They can be left as is.
   - But the columns from new_sp_m014 to newrel_f65 encode four variables in their names.
   
```{r}
#| label: WHO 1

who %>% 
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = c("diagnosis", "gender", "age"), 
    names_pattern = "new_?(.*)_(.)(.*)",
    values_to = "count"
  )

```
   

### Household: Multiple observations per row

The name of the column that belongs in the output is part of the column name in the input. 

```{r}
#| label: household

household

```


  - There are two values for each child: name and  dob (date of birth). 
  - These need to go into separate columns 
  - Again: multiple variables to names_to
       - use names_sep to split up each variable name
       - Note the special name .value: this tells pivot_longer() that that part of the column name specifies the “value” being measured 
       
```{r}
#| label: household 1

household %>% 
  pivot_longer(
    cols = !family, 
    names_to = c(".value", "child"), 
    names_sep = "_", 
    values_drop_na = TRUE
  )
  
  

```
      
       
       


## pivot_wider()

Makes a dataset wider by increasing the number of columns and decreasing the number of rows. 

It’s relatively rare to need pivot_wider() to make tidy data, but it’s often useful for creating summary tables for presentation, or data in a format needed by other tools.

### Capture-recapture data

The fish_encounters dataset, contributed by Myfanwy Johnston, describes when fish swimming down a river are detected by automatic monitoring stations:

```{r}
#| label: wider - fish

fish_encounters


```

To analyse this data, make each station a column.


```{r}
#| label: fish - columns

fish_encounters %>% 
  pivot_wider(
    names_from = station, 
    values_from = seen
  )

```


### Tidy census

The us_rent_income dataset contains information about median income and rent for each state in the US for 2017 (from the American Community Survey, retrieved with the tidycensus package).


```{r}
#| label: tidy census

us_rent_income

```

Here both estimate and moe are values columns, so we can supply them to values_from:


```{r}
#| label: census wider 

us_rent_income %>% 
  pivot_wider(
    names_from = variable, 
    values_from = c(estimate, moe)
  )

```










## Longer, then wider

For some problems, combining pivot_longer() and pivot_wider() is required.



### World bank
world_bank_pop contains data from the World Bank about population per country from 2000 to 2018.

```{r}
#| label: world bank 

world_bank_pop

```

 Start with the most obvious problem: year is spread across multiple columns.

```{r}
#| label: world bank year across multiple columns

pop2 <- world_bank_pop %>% 
  pivot_longer(
    cols = `2000`:`2017`, 
    names_to = "year", 
    values_to = "value"
  )
pop2



```

Now, consider the indicator variable:

```{r}
#| label: WB indicator

pop2 %>% 
  count(indicator)


```

Consider the sense of the data. 
  - SP.POP.GROW is population growth, 
  - SP.POP.TOTL is total population, 
  - and SP.URB.* are the same but only for urban areas. 
  
Split this up into two variables: 
  - area (total or urban) 
  - population or growth)


```{r}
#| label: WB 3

pop3 <- pop2 %>% 
  separate(indicator, c(NA, "area", "variable"))

pop3


```

Now, pivot(wider)

```{r}
#| label: WB pivot wider

pop3 %>% 
  pivot_wider(
    names_from = variable, 
    values_from = value
  )


```





